Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-03-07 manuscript/climate_sensitivity_draft.Rmd
Added citations for #78 and #77

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr', 'LeafArea'
), 'packages.bib')
```
\raggedright

@@

- We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree-ring record spanning historical droughts. Thus, these factors can be useful for predicting future drought responses under climate change. (**LS suggests take out "influence growth responses during drought, and". I agree this conclusion is a bit repetitive/could be shortened, but would target the "explain variation..." for removal**)

@@

Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [Dai et al. 2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015]. Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015]. Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts. However, it has proved difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but rarely consider the roles of tree size and microenvironment.

@@

Debates have also arisen regarding the traits influencing tree growth responses to drought. Differences have been observed between ring- and diffuse-porous species--with ring-porous species showing higher drought tolerance [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but this classification does not resolve differences among the many species within each category. Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011](**Martin-Benito & Pederson 2015, doi:10.1111/jbi.12462**) and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance ([Maréchaux et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), or the direction of response was not always consistent; for instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

@@

Topographic wetness index (TWI) was calculated using the dynatopmodel package in R (Figure S1) [@R-dynatopmodel]. Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006]. TWI calculation depends on an input of a digital elevation model (DEM; **Ian, specify resolution**), and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely. From our observations in the plot, TWI performed better at categorizing wet areas than the Euclidean distance from the stream.

@@

To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area. We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@noauthor_national_2018]. After filtering for missing and outlier values, the data were consolidated to represent the mean values per sensor height per day. The range of these means were then aggregated at a month scale.

@@

All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base]. All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively).

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.  The negative effect of height on $Rt$ held after accounting for species' traits. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality (**Sapes et al. 2019, doi: 10.1093/treephys/tpz062**).

@@

Our analysis has the limitation that canopy positions were recorded in 2018, as opposed to the years of the droughts. However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis. The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019]. It is safe to assume that currently suppressed trees been suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al. 2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*). The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species and associated traits at a single site [see also @elliott_forest_2015]. Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] (*Rosas et al. 2019, DOI: 10.1111/nph.15684*, [Fletcher et al. 2018](https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1164)), and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models (*Powell et al. 2016, DOI: 10.1111/gcb.13731*). Because both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic of traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (*Powell et al. 2016, DOI: 10.1111/gcb.13731*).

>>>>

```{r include=FALSE}
# automatically create a bib database for R packages
library(LeafArea)
knitr::write_bib(c(
  .packages(), 'knitr', 'rmarkdown', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'raster', 'elevatr', 'dynatopmodel', 'stringi', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'purrr', 'pointRes', 'dplR', 'data.table', 'tools', 'dplyr', 'reshape2', 'png', 'grid', 'gridExtra', 'cowplot', 'ggpubr', 'extrafont', 'rasterVis', 'tidyr', 'neonUtilities', 'lubridate', 'LeafArea'
), file = 'packages.bib')
```
\raggedright

@@

- We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree-ring record spanning historical droughts. Thus, these factors can be useful for predicting future drought responses under climate change. (**LS suggests take out "influence growth responses during drought, and". I agree this conclusion is a bit repetitive/could be shortened, but would target the "explain variation..." for removal** - *I agree with this*)

@@

Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [@dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015]. Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015]. Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts. However, it has proved difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but rarely consider the roles of tree size and microenvironment.

@@

Debates have also arisen regarding the traits influencing tree growth responses to drought. Differences have been observed between ring- and diffuse-porous species--with ring-porous species showing higher drought tolerance [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but this classification does not resolve differences among the many species within each category. Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent; for instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

@@

Topographic wetness index (TWI) was calculated using the dynatopmodel package in R (Figure S1) [@R-dynatopmodel]. Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006]. TWI calculation depends on an input of a digital elevation model (DEM; ~3.7 m resolution from the elevatr package [@R-elevatr]), and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely. From our observations in the plot, TWI performed better at categorizing wet areas than the Euclidean distance from the stream.

@@

To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area. We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018]. After filtering for missing and outlier values, the data were consolidated to represent the mean values per sensor height per day. The range of these means were then aggregated at a month scale.

@@

All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base]. All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively **krista just to remind you this TBD is here**).

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.  The negative effect of height on $Rt$ held after accounting for species' traits. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

@@

Our analysis has the limitation that canopy positions were recorded in 2018, as opposed to the years of the droughts. However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis. The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019]. It is safe to assume that currently suppressed trees been suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition [@sohn_potential_2016]. The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species and associated traits at a single site [see also @elliott_forest_2015]. Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015; @rosas_adjustments_2019; @fletcher_evolution_2018], and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017]. Because both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic of traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@powell_differences_2017].
